---
title: "Assignment.End-to-end"
output: html_document
date: "2023-05-04"
Author: Hanna Shrubshall 
---
##Assignment. ENd-to-end data analysis in R 
The following code is for the assignment using data from QFISH. 

*Note to self remember when copying path of file that the \ have to be changed to /

```{r}
install.packages("usethis")
library(usethis)
use_git_config(user.name = "hanna-shrubshall", user.email = "hanna.shrubshall@gmail.com")
Sharks <- read.csv("E:/github/MB5370-Module04/data/export.csv")
```

#Loading the required packages to create my plot 
```{r}
library(tidyverse) #this is for ggplot 
library(viridis) #this is for a differnet colour scheme 
library(hexbin)
```

#Cleaning up the data in R
- the following code is used to clean up the data so it is easier for R to read and easier to manipulate to place into a plot 
- Firstly you need to have Hmisc installed and downloaded into R
- Make a copy of the dataset 

```{r}
#install.packages("Hmisc")
library("Hmisc")

shark1 <- Sharks 
```

###Changing the labels/headers of the data 
```{r}
label(shark1) #this checks the labels of the data 
```

