---
title: "Workshop.3.Github"
output: html_document
date: "2023-05-04"
Author: Hanna Shurbshall 
---
#2.0 Workshop 1: Dataviv in R 

##2.1 Installing the tidyverse
```{r}
install.packages("tidyverse")
library(tidyverse)
```

##2.5 Obtain the data 
```{r}
mpg
```
##2.6 Create your first ggplot 
- ggplot() creates a coordinate system   that you can add layers to 
- the function "data =" is where you     input your dataset being used for the   plot
- geom_point() addds a layer of points   to your plots which creates a          scatterplot. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```
##2.7 Understand the 'grammar of graphics' 
###2.7.1 Graphing template 
- A general template for ggplot is to 'call ggplot, offer it data, then provide a geom function with a collection of mappings, which dictate how you plot will look' 
- The template is below 
```{r}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

###2.7.2 Aesthetic mappings 
- This refers to the size, shape and colour of data points 
- You need to attributes of a plot a geom and aesthetic 
Example, change point colour by class: 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```
Example, change point size by class: 
NOTE: Using size for a discrete variable is not advised. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Example, change point transparency by class (alpha):
- The alpha aesthetic is whats changing transparency 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```
Example, change point shape by class: 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Examples, things can also be set manually, like points colour: 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "purple")
```
Example question to try: What happens if you map an aesthetic to something other than a variable na,e, like aes(colour = displ <5)? 
- It's interesting to note that any number above 5 is coloured orange, where as any colour below 5 is coloured blue. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ <5))
```
##2.9 Facet and panel plots 
- This section allows you to develop separate plots for a range of reasons, most often to show a subset of data
- The function used for this is facet_wrap()
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))+ facet_wrap(~class,nrow = 2)
```
Example with more than one variable: 
- Function is facet_grid()
- Use ~ to separate variables 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))+ facet_grid(drv~cyl)
```
Example using a . to produce plots without rows or column dimensions: 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))+ facet_grid(.~cyl)
```
Exercise: 
What does nrow do? 
This dictates the number of rows there are 
what does ncol do? 
This dictates the number of columns there are 
What other options control the layout of the individual panels? 
The function as.table lays the facets out as a table if TRUE with the highest values at the bottom right. if FALSE the facets are laid out like a plot with the highest value at the top right. 

##2.10 Fitting simple lines
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) 
```
Example, display the same data as a smooth line fir through the points using geom_smooth()
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
```{r}
ggplot(data = mpg) + 
#  geom_point(mapping = aes(x = displ, y = hwy))#points horrible
  geom_smooth(mapping = aes(x = displ, y = hwy))#try mooth line
```

Example, changing the line type: 
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

Example, set a group aesthetic to a categorical variable to draw multiple objects: 
- using a group argument to show that the data is in fact grouped - it won't add a new legend but can easily be used to vary your plots by discrete variables. 
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```
Example, change the colour of each line based on drv value: 
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, colour = drv), show.legend = FALSE)
```

Example, plotting multiple geoms on the single plot: 
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(x = displ, y = hwy))
```

Example, creating the same plot as above, BUT in a more effcient manner: 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point()+
  geom_smooth()
```

Example, using mappings in specific layers to enable the display of different aesthetics in different layers: 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(color = class))+
  geom_smooth()
```

Example, specify different data for each layer, here a filter is used to select a subset of data and plot only that subset (class = "subcompact"): 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(mapping = aes(color = class))+
  geom_smooth(data = filter(mpg, class =="subcompact"), se = FALSE)
```

Exercise:
1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
- geom_smooth 
- geom_boxplot
- geom_histogram
- geom_area 

2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
- it will produce points and line data 

3. Will these two graphs look different? Why/why not?
These two graphs will look the same because the first set of cod is  the more efficient way to write the second lot of code. Hence, they will produce the same map. 

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```
##2.11 Transformations and Stats
The following is going to be using a different data set 
- The diamonds dataset comes in ggplot2 and contains information about ~54,000 diamonds, including the price, carat, color, clarity and cut of each diamond 

###2.11.1 Plotting statistics
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

- bar charts, histograms, and frequency polygons bin your data and then plot bin counts, the number of points that fall in each bin.
- smoothers fit a model to your data and then plot predictions from the model.
- boxplots compute a robust summary of the distribution and then display a specially formatted box.

The algorithm used to calculate new values for a graph is called a stat, short for statistical transformation. 

Example, using geoms and stats interchangeably: 
```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```
###2.11.2 Overriding defaults in ggpplot2

Example, understanding how to override a default stat: 
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
Example, overriding a default mapping from transformed variables to aesthetics (displaying a bar chart of proportion of your total diamond data set rather than a count): 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop), group = 1))
```
###2.11.3 Plotting statistical details 
Example, understanding more about these transformations in your plot: using stat_summary()
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```
Example, you need to set the group = x number, otherwise the plots data will be represented as the same. 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop)))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))
```
##2.12 Positional adjustments 
Example using colour or fill to change aspects of bar colours 
- The first graph has coloured borders where as the second is filled in 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
Example, using these aesthetics to colour by clarity: 
- the stacking is done automatically 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```
Position adjustments allow you to customise plots in 3 ways: 
1. identity (raw data)
2. fill (changes in heights)
3. dodge (forces ggplot2 to not put things on top of each other)

- position = "identity" means you can place each object exactly where it falls in the context of the graph (good for scatter plots)
```{r}
#To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

#To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```
Example, position = "fill" makes each stacked bar the same height: 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```
Example, position = "dodge" places overlapping objects directly BESIDE one another 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```
Example, position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap (useful for scatterplots, not bar graphs)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```
##2.13 The layered grammar of graphics 
This is the template for creating ggplots 
```{r}
ggplot(data = <DATA>)+ 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>)),
     stat = (<STAT>), 
     position = (<POSITION>) +
  <FACET_FUNCTION>
 
```
----------------------------------------------------

##Workshop 3: Ussing ggplot2 for communication 
##3.1 Labels 
Example, adding labels to ggplots are important for communicating what you are plotting. This eaxmple is inputting a title:
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

Example, adding more text: 
- subtitle adds additional detail in a smaller font beneath the title and caption adds text at the bottom right of the plot 
- caption adds text at the bottom right of the plot, often used to describe the source of the
data.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

Example using labs() to replace axis label and legend titles: 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```

##3.2 Annotations 
Adding text to the plot itself. 
- geom_text() to ad textual labels to plots 
- geom_point(), rather than a shape geometry it can add a label. 
- the "nudge()" function can be used to mave certain text around so that it isn't overlapping 

Example, adding a label that calls in the values from the data frame: 
```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
```

##3.3 Scales 
ggplot2 automatically adds scales for you as seen below: 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))
```
To tweak the scales you can do it by offering values to the scale parameters: 
- if you just add a number in the scale_x_continuous() brackets it only changes the label. But if you add in the breaks=seq then you can change the scale 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous(breaks = seq(0, 10, by = 1)) +
  scale_y_continuous(breaks = seq(15, 50, by = 5)) +
  scale_colour_discrete(10)
```
##3.4 Axis ticks 
Example changing the ticks on the axes: 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```
##3.5 legens and colour schemes 
Example, changing the position of the legend using theme(): 
- Can also use legend.position = "none" to supress the display of the legend altogether 
```{r}
base <- ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
```

```{r}
base <- ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(colour = class))
base + theme(legend.position = "top")#can use this and change the position as a way to change the location of the legend
```
##3.6 Replacing a scale 
There are two types of scales that are most likely going to be used: contiuous position scales and colour sclaes 

Example, using logtransform to easily visualise data: 
```{r}
ggplot(diamonds, aes(carat, price))+
  geom_bin2d()+ 
  scale_x_log10(5)+ 
  scale_y_log10(10)
```

Example, changing the colour scale: 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```

Example, adding a redundant shape mapping (this helps to ensure the plot is interpretable in black and white): 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```

Colours  http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/ 
- the above website is for more colour options off colorbrewer 

When there are predefined colours that you want to use the function scale_colour_manual() can be used:

```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```

Another colour scheme used is the viridis colour scheme: scale_colour_viridis()
- downloading the viridis package 
```{r}
install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)
```

```{r}
df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

##3.7 Themes 
Customising the entire theme of your plot 
- ggplot has 8 themes by default 
Example, the below graph is the BLACK AND WHITE, dark on light ggplot2 theme 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

Example, the below graph has LIGHT GRAY lines and axes, to direct more attention to the data. 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()
```
Example, the graph below is a CLASSIC theme with the x and y axis lines and no gridlines: 

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()
```

Example, the graph below is the DARK THEME  
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```

Example, you can develop your own themes by setting more arguments: 
- This is an example of one that has been tailored. 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial")) 
```

##3.8 Saving and exporting your plots 
There are two main ways to export your plots out of R: 
1. Using ggsave()
2. Or export with your code ising R markdown via knitr
```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
```

----------------------------------------------------------------
##4.0 Workshop 3: Reproducible work  
Below the code is installing Git
```{r}
install.packages("usethis")
library(usethis)
use_git_config(user.name = "hanna-shrubshall", user.email = "hanna.shrubshall@gmail.com")
```

##4.9.1 Pull
- Pulling a repository simply downloads it from the remote server onto the computer. 
##4.9.2 Stage 
Git stage prepares the file for committing it. 
- The files in the Git tab need to be ticked as standard, meaning you want the files to be sent to your remote repo when you go ahead and commit. 

##4.9.3 Commit 
This is the function that saves this 'version' of your repo into your repository's version history. 
-At this point you can write readable messages of what you've done to your repo. 

##4.9.4 Push 
This is the function that pushes your changes to the remote server 
-It's best practice to commit and push your code at the same time

------------------------------------------------------------------------
##5.0 Workshop 4: Data Wrangling in R
- A tibble is a data frame that is slightly adjusted to help users keep up with the evolving applications of R. 
- Tibbles are different to R's built-in data frame which will be called data.frame S. 
- The tibble package is part of the tidyverse package 

##5.3 WHat is a tibble? 

```{r}
library("tidyverse")
```
- R packages use regular data frames. To turn said regular data frames into tibbles you can use the function as_tibble(). 

```{r}
as_tibble(iris)
```

- You can also create a new tibble from individual vectors using tibble(). This will automatically recycle inputs of length 1, and allows you to refer to variables that you just created. 

*NOTE, tibble() does not change input types, names of variables or create row names
Example below: 

```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

Tibbles can have non-syntactic names (which aren't valied in R).
- For example names with spaces or containing special characters
If there are non-syntatic variables in your tibble, you'll need to refer to them  with backsticks. 
*NOTE, you'll need these backsticks for these variables when using other packages such as ggplot2, dplyr, tidyr, etc. 
Example below: 

```{r}
tb <- tibble(
  `:)` = "smile",
  ` ` = "space", 
  `2000` = "number"
)
tb

```

Another way to create a tibble is with the function tribble() (meaning transposed tibble). 
- This is a customised way for data entry in code: column headings are defined by formulas, and entries are separated by commas
Example below: 

```{r}
tribble(
   ~x, ~y, ~z,
  #Header
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

What is the difference between tibbles and data.frame? 
- Tibbles print the first 10 rows and all columns that can fit on screen 
- Tibbles also print the type of each data next to its name (character, string, numeric, etc). 
- Tibbles help your console from getting overwhelmed by printing massive data frames. 

```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

However, sometimes you need to alter the default display. 
- Print() can be used to designate the number of ros (n) and display width (width = Inf #displays every column). 

Example below: 

```{r}
install.packages("nycflights13")
library("nycflights13")
nycflights13::flights %>% 
  print(n = 10, width = Inf)
```

You can also control the default print behaviour by setting options: 
- options(tibble.print_max = n, tibble.print_min = m): if more then n rows, print only m rows
- Use options(tibble.print_min = Inf) to always show all rows.
-Use options(tibble.width = Inf) to always print all columns, regardless of the width of the screen.
-Use R’s built-in viewer with View()
-You can see a complete list of options by looking at the package help with package?tibble.

SOme tools are needed for pulling out single variables from our data frame. [[ extracts by name or position and $ extracts only by name. 

Example below: 

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x

df[["x"]]

# Extract by position
df[[1]]

```

Need to use a . as a place holder if we want to use these tools in a pipe (%>%). 
Example below: 

```{r}
df %>% .$x
df %>% .[["x"]]
```

Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

```{r}
df <- tibble(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```


##5.4 How can I import data? 
- read_csv() reads comma delimited files
- read_csv2() reads semicolon separated files 
- read_tsv() reads tab delimited files 
- read_delim() reads in files with any delimiter
- read_fwf() reads fixed width files. You can specify fields with their widths with either fwf_widths() or position with fwf_positions()
- read_table reads a common variation of fixed width files where columns are separated by white space. 
- read_log() reads Apache style log files 

Most important argument to read_csv() is the file path.
- when you run read_csv() it prints out a column specification that gives the name and type of each column 

```{r}
read_csv("a,b,c
1,2,3
4,5,6")
```
*NOTE read_csv() uses the first line of data as the column names 

You may not want the first line of data to be the column names hence you might want to skip these lines. 
- Use function skip = n (where n is the number of lines you want to skip over). 
- Alternativey, use comment = '#' it will drop all lines starting with a # or whatever character you designate 

```{r}
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)

read_csv("# A comment I want to skip
  x,y,z
  1,2,3", comment = "#")
```

Another reason to adjust the read_csv() default is if your data does not have column names. 
- Use col_names = FALSE (this tells read_csv() not to treat the first row as headings, but to label them sequentially instead from X1 to Xn)

*NOTE \n is a shortcut to add a new line

Example below: 

```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE)
```

You can also pass col_names a character vector to be used as the column names:

```{r}
read_csv("1,2,3\n4,5,6", col_names = c("Bob", "Charlie", "Sam"))
```

Another option that commonly needs tweaking is na: this specifies the value (or values) that are used to represent missing values in your file:

```{r}
read_csv("a,b,c\n1,2,.", na = ".")
```

##5.5 Tidying data using Tidyr
```{r}
library(tidyverse)
```

###5.5.1 Tiday data 
There are many ways to display a given data set, but not every way is easy to use for analysis. Note that in the example below, only table1 is “tidy”.

```{r}
table1

table2

table3

table4a

table4b
```

How we make our dataset a tidy dataset is by following three interrelated rules. 
1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

 Here are some examples of how you might work with tidy table1 from the previous example.	
 
```{r}
table1 %>% 
  mutate(rate = cases / population * 10000)

table1 %>% 
  count(year, wt = cases)

library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```
 
###5.5.2 Spreading and gathering data tables 
Common problems with untidy data: 
1. One variable is spread across multiple columns
2. One observation is scattered across multiple rows
- to fix said problems the following to functions will be used 
1. pivot_longer() A common problem is a dataset has some column names that are not the names of the variables, but the values of the variables

Example below: 

```{r}
table4a
```

To tidy this type of dataset: 
- The set of columns whose names are values, not variables. In this example, those are the columns 1999 and 2000.
- The name of the variable to move the column names to. Here it is year.
- The name of the variable to move the column values to. Here it’s cases.

Example of using picot_longer()

```{r}
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```

```{r}
table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```
COmbinning the tidied versions of table 4a + 4b into a single tibble 

```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
left_join(tidy4a, tidy4b)
```

2. pivot_wider()
```{r}
table2

table2 %>%
    pivot_wider(names_from = type, values_from = count)
```

###5.5.3 Seperating and uniting data tables 
In table 3, there is one column (rate) that has two variables (cases & population). To address this you can use the function separate(), which seperates one column into multiple columns. 

Example below: 
```{r}
table3

table3 %>% 
  separate(rate, into = c("cases", "population"))#this separates the data into cases and population 


table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)#Both cases and population are listed as character types. This is a default, the values in these columns are actually numbers so we need to convert them to better types by using convert=TRUE


table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
#Can also pass a vector of integers to sep. separate() will interpret the integers as positions to split at. 


```
Can perform the inverse of separate by using unite() to combine multiple columns into a single column. 
- unite() takes a data frame, the name of the new variable and a set of columns to combine using dplyr::select(). 

```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```
###5.5.4 Handling missing values 

Missing values are either blank or generally filled by NA. 
Example, below has two points that have missing values: 

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

*NOTE, an "NA" (explicit absence) indicates the presence of absent data. A "blank cell" just indicates the absence of data (implicit absence). 

Example, below making implicit values explicit: 

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return)
```

The explicit missing claues may not be important in other representations of the data, hence you can set values_drop_na=TRUE in pivot_longer() to turn explicit values implicit

Example, below: 
```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(
    cols = c(`2015`, `2016`), 
    names_to = "year", 
    values_to = "return", 
    values_drop_na = TRUE
  )
```

Another function that can be used is the complete() function. This taks a set of columns and finds all the unique combinations and then ensures the original dataset contains all of those values, including filling in explicit NA where needed. 

```{r}
stocks

stocks %>% 
  complete(year, qtr)
```

Similarly, the fill() function can be used to fill in the missing values that were meant to be carried forward. 

Example, below: 

```{r}
 treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment

treatment %>% 
  fill(person)
```

##5.6 Learning relational data
Relations are defined between a pair of tables. All other relations are built up from this simple idea: the relations of three or more tables are always a property of the relations between each pair. Sometimes both elements of a pair can be the same table! This is needed if, for example, you have a table of people, and each person has a reference to their parents.

```{r}
library(tidyverse)
library(nycflights13)
```

*NOTE, nycflights13 contains 4 tibbles that are related to flights. 

```{r}
airlines #airlines lets you look up the full carrier name from its abbreviated code

airports #airports give information about each airport, identified by the faa airport code

planes #gives information about each plane, identified by its tailnum

weather #gives the weather at each NYC airport for each hour 


```

###5.6.1 Joining datasets 
To join datasets together you need to identify the keys.
- A key is a variable (or set of variables) that uniquely identifies an observation. 

- There are two types of keys
  1. Primary key, uniquely identifies an observation in its own table
  2. Foreign key, uniquely identifies an observation in another table
  
- Once the primary keys have been identified in your table, its good practice to verify that they do indeed uniquely identify each observation. Using the function count(). 

*NOTE, if a table lacks a primary key, you can add one with mutate() and row_number()

```{r}
#Tables 
planes %>% 
  count(tailnum) %>% 
  filter(n > 1)

weather %>% 
  count(year, month, day, hour, origin) %>% 
  filter(n > 1)

flights %>% 
  count(year, month, day, flight) %>% 
  filter(n > 1)

#the key is tailnumber not flights
flights %>% 
  count(year, month, day, tailnum) %>% 
  filter(n > 1)
```

###5.6.2 Mutating joins 
Mutating a join allows you to combine variables from two tables. 
- it first matches variables by their keys, then copies across variables from one table to the other. 

```{r}
#Narrow subset of data 
flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)

#Use of left_join function 
flights2 %>%
  select(-origin, -dest) %>% 
  left_join(airlines, by = "carrier")

#Use of mutate function (creates the same result as previous code)
flights2 %>%
  select(-origin, -dest) %>% 
  mutate(name = airlines$name[match(carrier, airlines$carrier)])

#Dive into how mutating joins work in detail
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)


#output of an inner join containing the key, x & y values
x %>% 
  inner_join(y, by = "key")


#outer join keeps observations that appear in at least one of the tables, there are 3 types: left_join(keeps all observations in x), right_join(keeps all observations in y), full_join(keeps all observations in x and y)
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     1, "x4"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2"
)
left_join(x, y, by = "key")

#Both tables have duplicate keys, which is usually an error. When you join duplicated keys, you get all possible combinations = the Cartesian product: 
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     2, "x3",
     3, "x4"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     2, "y3",
     3, "y4"
)
left_join(x, y, by = "key")

#Using the function by to join tables 
flights2 %>% 
  left_join(weather)

flights2 %>% 
  left_join(planes, by = "tailnum")


#specifying what point you want to join 
flights2 %>% 
  left_join(airports, c("dest" = "faa"))

flights2 %>% 
  left_join(airports, c("origin" = "faa"))


```

###5.6.3 Filtering Joins 
```{r}

```

